---
title: "Use case 1: Effect of gut EVs in Kupffer cells (Tabula Muris)"
author: "Estefania"
date: "2025-03-10"
output: md_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path='Figs/', message=FALSE)
```

In this fuse case we will explore the effect of the proteomic EV cargo, isolated from the gut of a prediabetic mouse model, in upregulated Kupffer cells (when in comparison with hepatocytes). 

0. Load needed packages
```{r}

library("nichenetr")
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(dplyr)
library(here)
```

1. Load filtered Seurat object
The dataset originates from the liver (FACS) subset of the Tabula Muris Atlas (Tabula Muris Consortium, 2018). To ensure data quality and relevance, we filtered the Seurat object to retain only features with at least 10 counts, present in at least 5% of either Kupffer cells or hepatocytes populations. 

```{r}
new_seurat_object <- readRDS(url("https://zenodo.org/record/15014486/files/filtered_seurat_obj_use_case_1.rds"))
```

2. Quality control new Seurat object with filtered features

Exploring subset

```{r}
new_seurat_object@meta.data$cell_ontology_class %>% table() 
```

Visualize QC metrics as a violin plot

```{r}
VlnPlot(new_seurat_object, features = c("nFeature_RNA", "nCount_RNA"), ncol = 3)
```

Visualize cell populations in a UMAP

```{r}
DimPlot(new_seurat_object, reduction = "umap")

```

3. Subsetting cell populations of interest (Kupffer cells and hepatocytes)
We want to keep only Kupffer cells and hepatocytes cell populations

```{r}
seurat_obj <- subset(new_seurat_object,
        subset = cell_ontology_class %in% c("Kupffer cell", "hepatocyte"))

Idents(object = seurat_obj) <- seurat_obj@meta.data$cell_ontology_class
```

4. Exploring subset

```{r}
seurat_obj@meta.data$cell_ontology_class %>% table() 
```

5. Loading Nichenet networks from Zenodo

```{r}
organism <- "mouse"

if(organism == "human"){
  lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
  weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final.rds"))
} else if(organism == "mouse"){
  lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_mouse_21122021.rds"))
  weighted_networks <- readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final_mouse.rds"))

}

lr_network <- lr_network %>% distinct(from, to)
head(lr_network)
```
6. Download the EV-Net EV_cargo_target_matrix from https://zenodo.org/records/15019664/files/EV_cargo_target_matrix.rds and load it into your R environment.

```{r}
EV_cargo_target_matrix <- readRDS("~/embo/Netiev/EV_cargo_target_matrix.rds")
#Replace file path with the path in which you stored the EV_cargo_target_matrix rds file
```


7. Define receiver cell populations (Kupffer cells and hepatocytes) and get features present in both
```{r}
# Step 1: Extract the cells belonging to each cell type
cells_type_1 <- WhichCells(new_seurat_object, idents = "Kupffer cell")
cells_type_2 <- WhichCells(new_seurat_object, idents = "hepatocyte")

# Step 2: Subset the Seurat object to get the count matrix for each cell type
data_matrix <- GetAssayData(new_seurat_object, slot = "counts")

# Get the features (genes) expressed in each cell type
# Features expressed in at least one cell for cell_type_1
features_type_1 <- rownames(data_matrix)[rowSums(data_matrix[, cells_type_1] > 0) > 0]

# Features expressed in at least one cell for cell_type_2
features_type_2 <- rownames(data_matrix)[rowSums(data_matrix[, cells_type_2] > 0) > 0]

# Step 3: Combine the two lists of features and remove duplicates
combined_features <- unique(c(features_type_1, features_type_2))

```


```{r}
receiver = c("Kupffer cell", "hepatocyte")

expressed_genes_receiver <- combined_features
```

8. Defining expressed_interactors,  as all proteins expressed in the receiving cell populations (eceptors, downstream proteins and transcription factors) which could interact with the potential_EV_cargo

```{r}
all_genes <- unique(rownames(EV_cargo_target_matrix))  
expressed_interactors <- intersect(all_genes, expressed_genes_receiver)

lr_sig <- weighted_networks[["lr_sig"]]
gr <- weighted_networks[["gr"]]

potential_EV_cargo_prot <- lr_sig[lr_sig$to %in% expressed_interactors, "from"]
potential_EV_cargo_tf <- gr[gr$to %in% expressed_interactors, "from"]
potential_EV_cargo <- unique(c(potential_EV_cargo_prot$from, potential_EV_cargo_tf$from))
```


9. Load the differentially abundant EV_cargo present in the predeiabetic condition (proteins with a p value < 0.05 and a negative logFC following an empirical Bayes t-test using limma, comparing healthy vs. prediabetic conditions)

```{r}
prediabetic_EV_cargo <- readRDS("~/embo/Netiev/prediabetic_EV_cargo.rds")
#This dataset will be in the package
```

10. Get list of EV_cargo

```{r}
EV_cargo <- intersect(potential_EV_cargo, prediabetic_EV_cargo)
```

11. The EV_cargo_target_matrix is quite heavy in size (3.6 gb). If you have limited ram memory, we recommend you to run this chunk. It will keep only the expressed_genes_receiver and potential_EV_cargo in the matrix, reducing its size. If you are working with 32gb of ram or more, this step is optional.

```{r}
EV_cargo_target_matrix <- EV_cargo_target_matrix[rownames(EV_cargo_target_matrix) %in% expressed_genes_receiver, colnames(EV_cargo_target_matrix) %in% potential_EV_cargo]

```

12. Define the gene set of interest
Note: We had to add an identity class label with Idents

```{r}
condition_oi <-  "Kupffer cell"
condition_reference <- "hepatocyte"

seurat_obj_receiver <- subset(seurat_obj, idents = receiver)

DE_table_receiver <-  FindMarkers(object = seurat_obj,
                                  ident.1 = condition_oi, ident.2 = condition_reference,
                                  group.by = "cell_ontology_class",
                                  min.pct = 0.05) %>% rownames_to_column("gene")

geneset_oi <- DE_table_receiver %>% filter(p_val_adj <= 0.05 & avg_log2FC >= 0.25) %>% pull(gene)
geneset_oi <- geneset_oi %>% .[. %in% rownames(EV_cargo_target_matrix)]
```

13. Defining background genes

```{r}
background_expressed_genes <- expressed_genes_receiver %>% .[. %in% rownames(EV_cargo_target_matrix)]
```

14. Perform EV cargo activity analysis

```{r}
EV_cargo_activities <- predict_ligand_activities(geneset = geneset_oi,
                                               background_expressed_genes = background_expressed_genes,
                                               ligand_target_matrix = EV_cargo_target_matrix,
                                               potential_ligands = EV_cargo)

EV_cargo_activities <- EV_cargo_activities %>% arrange(-aupr_corrected) %>% mutate(rank = rank(desc(aupr_corrected)))
EV_cargo_activities

```
15. Visualization of top-ranked EV cargo

```{r}
p_hist_EV_cargo_activity <- ggplot(EV_cargo_activities, aes(x=aupr_corrected)) + 
  geom_histogram(color="black", fill="darkorange")  + 
  geom_vline(aes(xintercept=min(EV_cargo_activities %>% top_n(30, aupr_corrected) %>% pull(aupr_corrected))),
             color="red", linetype="dashed", size=1) + 
  labs(x="EV cargo activity (PCC)", y = "# EV cargo") +
  theme_classic()

p_hist_EV_cargo_activity
```
16. We can also visualize the EV cargo activity measure (AUPR) of these top-ranked EV cargo

```{r}
best_upstream_EV_cargo <- EV_cargo_activities %>% top_n(30, aupr_corrected) %>% arrange(-aupr_corrected) %>% pull(test_ligand)

vis_EV_cargo_aupr <- EV_cargo_activities %>% filter(test_ligand %in% best_upstream_EV_cargo) %>%
  column_to_rownames("test_ligand") %>% select(aupr_corrected) %>% arrange(aupr_corrected) %>% as.matrix(ncol = 1)

(make_heatmap_ggplot(vis_EV_cargo_aupr,
                     "Prioritized EV cargo", "EV cargo activity", 
                     legend_title = "AUPR", color = "darkorange") + 
    theme(axis.text.x.top = element_blank()))  
```

17. Infer target genes  of top-ranked EV cargo

```{r}
active_EV_cargo_target_links_df <- best_upstream_EV_cargo %>%
  lapply(get_weighted_ligand_target_links,
         geneset = geneset_oi,
         ligand_target_matrix = EV_cargo_target_matrix,
         n = 80) %>%
  bind_rows() %>% drop_na()
```

```{r}
active_EV_cargo_target_links <- prepare_ligand_target_visualization(
  ligand_target_df = active_EV_cargo_target_links_df,
  ligand_target_matrix = EV_cargo_target_matrix,
  cutoff = 0.5) 
```

```{r}
order_EV_cargo <- intersect(best_upstream_EV_cargo, colnames(active_EV_cargo_target_links)) %>% rev()
order_targets <- active_EV_cargo_target_links_df$target %>% unique() %>% intersect(rownames(active_EV_cargo_target_links))

vis_EV_cargo_target <- t(active_EV_cargo_target_links[order_targets,order_EV_cargo])

target_genes_heatmap <- make_heatmap_ggplot(vis_EV_cargo_target, "Prioritized EV cargo", "Predicted target genes",
                    color = "purple", legend_title = "Regulatory potential") +
  scale_fill_gradient2(low = "whitesmoke",  high = "purple")
```

```{r}
target_genes_heatmap
```
18. Save plot (optional)
```{r}
png("target_genes_heatmap_prediabetic_EV_cargo_target_matrix.png", res = 300, width = 5000, height = 2000)
print(target_genes_heatmap)
```

Next we will build an interaction network between one of the top-ranked EV cargo: the Scly protein, and two of its targets, with biological relevance for prediabetes: Zeb2 and Foxo1 (choice was made after studying the target_genes_heatmap). 

19. Loading Nichenet's sig_network and gr_network from zenodo

```{r}
sig_network <- readRDS(url("https://zenodo.org/records/7074291/files/signaling_network_mouse_21122021.rds"))
gr_network <- readRDS(url("https://zenodo.org/records/7074291/files/gr_network_mouse_21122021.rds"))
```

20. Inferring ligand-to-target signaling paths NCD

```{r}
EV_cargo_oi <- "Scly"
targets_oi <- c("Zeb2", "Foxo1")

active_signaling_network <- get_ligand_signaling_path(ligands_all = EV_cargo_oi,
                                                      targets_all = targets_oi, 
                                                      weighted_networks = weighted_networks,
                                                      ligand_tf_matrix = EV_cargo_target_matrix,
                                                      top_n_regulators = 4,
                                                      minmax_scaling = TRUE) 


graph_min_max <- diagrammer_format_signaling_graph(signaling_graph_list = active_signaling_network,
                                                   ligands_all = EV_cargo_oi, targets_all = targets_oi,
                                                   sig_color = "indianred", gr_color = "steelblue")

# To render the graph in RStudio Viewer, uncomment following line of code
 #DiagrammeR::render_graph(graph_min_max, layout = "tree")

# To export/draw the svg, you need to install DiagrammeRsvg
#graph_svg <- DiagrammeRsvg::export_svg(DiagrammeR::render_graph(graph_min_max, layout = "tree", output = "graph"))
#cowplot::ggdraw() + cowplot::draw_image(charToRaw(graph_svg))
```

21. Adding annotations of the collected data sources support the interactions in this network

```{r}
data_source_network <- infer_supporting_datasources(signaling_graph_list = active_signaling_network,
                                                    lr_network = lr_network, sig_network = sig_network, gr_network = gr_network)
head(data_source_network) 
```

22. Exporting interaction network to cytoscape

```{r}
output_path <- "~/embo/Netiev/"
write_output <- TRUE # change to TRUE for writing output

# weighted networks ('import network' in Cytoscape)
if(write_output){
  bind_rows(active_signaling_network$sig %>% mutate(layer = "signaling"),
            active_signaling_network$gr %>% mutate(layer = "regulatory")) %>%
    write_tsv(paste0(output_path,"weighted_signaling_network_prediabetic_Scly.txt")) 
}

# networks with information of supporting data sources ('import network' in Cytoscape)
if(write_output){
  data_source_network %>% write_tsv(paste0(output_path,"data_source_network_prediabetic_Scly.txt"))
}

# Node annotation table ('import table' in Cytoscape)
specific_annotation_tbl <- bind_rows(
  tibble(gene = EV_cargo_oi, annotation = "EV cargo"),
  tibble(gene = targets_oi, annotation = "target"),
  tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(c(targets_oi,EV_cargo_oi)) %>% intersect(lr_network$to %>% unique()), annotation = "receptor"),
  tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(c(targets_oi,EV_cargo_oi)) %>% intersect(gr_network$from %>% unique()) %>% setdiff(c(data_source_network$from, data_source_network$to) %>% unique() %>% intersect(lr_network$to %>% unique())),annotation = "transcriptional regulator")
)
non_specific_annotation_tbl <- tibble(gene = c(data_source_network$from, data_source_network$to) %>% unique() %>% setdiff(specific_annotation_tbl$gene), annotation = "signaling mediator")

if(write_output){
  bind_rows(specific_annotation_tbl, non_specific_annotation_tbl) %>%
    write_tsv(paste0(output_path,"annotation_table_prediabetic_Scly.txt"))
}
```
